---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.** 26 departments have offered the most courses with "Climate" in the title. Atmospheric and Oceanic Science has offered the most, with 68.

```{r}
umd_courses |>
  group_by (department) |>
  filter (str_detect(title, "Climate") & seats > 0) |>
  summarise (classes = n()) |>
  arrange (desc(classes))
```

**Q2.** Of the departments listed in A2, which one is the most interesting to you, and why?
**A2.** Journalism has two classes with "climate" in the title. As a journalism major, I am curious to see what the connection is there.


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.** Business and Management has the most courses with at least one syllabus posted in the Fall 2023 term. The top five departments here include departments in the business schools, arts, and sciences.

```{r}
umd_courses |>
  group_by (department) |>
  filter (str_detect (term, '202308') & syllabus_count > 0) |>
  summarise (courses = n()) |>
  arrange (desc(courses))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** Wes Moore's committee spending generally increased over time. The lowest spending in any month in 2022 was still greater than the highest spending in any month in 2021.

```{r}
  maryland_expenses |>
  mutate(expenditure_date=mdy(expenditure_date)) |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  filter (committee_name == 'Moore  Wes For Maryland') |>
  group_by(month) |>
  summarise (total_amount = sum(amount)) |>
  arrange (desc(total_amount))
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5**  R created a table with two columns. The most money was spent in media, with $30,256,356.74.

```{r}
maryland_expenses |>
  mutate(
    spending_location = case_when(
      str_detect(address, "Maryland") ~ "In-state" ,
      str_detect(address, " California ") ~ "Out of state",
        str_detect(address, " Washington ") ~ "Out of state",
        str_detect(address, " Louisiana ") ~ "Out of state",
        str_detect(address, " Florida ") ~ "Out of state",
        str_detect(address, " North Carolina ") ~ "Out of state",
        str_detect(address, " Massachusetts ") ~ "Out of state",
        str_detect(address, " West Virginia ") ~ "Out of state",
        str_detect(address, " Virginia ") ~ "Out of state",
        .default = NA
    )) |>
    group_by(expense_category) |>
    summarise (total_amount = sum(amount)) |>
  arrange (desc(total_amount))
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.** Q: Which department has the most courses in the fall 2023 term with at least 10 available seats?
A: Business and Management has 151 courses in fall 2023 with at least 10 available seats, the most of any department.

```{r}
umd_courses |>
  group_by (department) |>
  filter (str_detect (term, '202308') & seats > 9) |>
  summarise (courses = n()) |>
  arrange (desc(courses))
```
