---
title: "lab_01"
author: "Derek Willis"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you?
**A1** Engineering, Cooperative Education is the department with the highest mean number of seats. I find this surprising because I know a lot of engineering majors so it's hard to believe there are that many open seats in engineering classes.

```{r}
umd_courses |>
     group_by(department) |>
    summarise(
        mean_seats = mean(seats)
    ) |>
arrange (desc(mean_seats))

```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first. 
**A2** The greatest number of times any individual class has been offered in our dataset is 22.

```{r}
umd_courses |>
  group_by(id) |>
  summarise(
    count_id = n()
  ) |>
arrange (desc(count_id))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? **A3** Keith Yearwood has taught 61 classes in Geographical Sciences, which is the most of any instructor in one department.

```{r}
umd_courses |>
  group_by(instructors, department) |>
  summarise(
    count_instructors = n()
  ) |>
  arrange(desc(count_instructors))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4** If I could filter out the N/A's, it would be much easier to find which instructor has taught the most classes. I am still confident that my answer is correct because I went page by page and found the first name, as I have the data sorted by descending order of classes taught.

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5** My question: What is the average number of classes offered per department? - This question is complicated, as I have to first group by both department and term. I would have to calculate the mean number of classes offered for each individual department for each term, and then calculate the mean of all of those numbers. We are not given the number of classes for each department in the dataset, however, so I'm not sure what code to use to answer this question. I imagine if it can be done it's probably pretty complicated, as I was not able to figure it out.

```{r}
umd_courses |>
  group_by(department, term) |>
  summarise(
    mean_
  ) 
```
