---
title: "reverse_engineering_project"
author: "Hayden Sweeney"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

## Load libraries.

Load the required libraries for analyzing the data.

```{r}
options(scipen = 999)
library(tidyverse)
library(janitor)
library(readr)
library(lubridate)
```

## Loading and cleaning data

Load the data we will use for this analysis, and change the amount column from a character to a number.

```{r}
sports_owner_donations <- read_csv("sports-political-donations.csv") |>
  mutate(Amount = str_replace_all(Amount,'\\$','')) |>
  mutate(Amount = str_replace_all(Amount, '\\,','')) |>
  mutate(Amount = as.numeric(Amount)) |>
  clean_names()
```

In this notebook, we are reverse engineering what we believe to be the five most important findings from this story.

## 1: ""Of the 160 owners and commissioners in ESPN’s database, 74 contributed a majority of their publicly searchable funds to Republican campaigns, with 48 donating a majority to Democrats and the rest sending most of their money to causes that aren’t fully aligned with either party."

## How many owners contributed donations to the Republican party?

Our answer is 104. This is different from the 74 number above likely because some of these owners onoy donated very little to the Republican party, and the majority to a different party or likely bipartisan. 

```{r}
sports_owner_donations |>
  filter(str_detect(party, 'Republican')) |>
  group_by(owner) |> 
  summarise(total = n())
```


## 2: "“American professional sports owners have contributed nearly $47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight, including $10 million to Republican causes and $1.9 million to Democratic causes so far in the 2020 election cycle.” 

## Find the total of all donations from all owners. 

The total amount that all owners have donated since 2015 is $46,978,697. 

```{r}
sports_owner_donations |>
  summarise(sum(amount))
```

## Find how much money owners donated to Republicans in the 2020 election cycle.

Owners donated $10,111,284 to Republicans in the 2020 election cycle.

```{r}
sports_owner_donations |>
  filter(str_detect(party, 'Republican')) |>
  filter(election_year == 2020) |>
  summarise(sum(amount))
```
## Find how much money owners donated to Democrats in the 2020 election cycle.

Owners donated $1,877,133 to Democrats in the 2020 election cycle.

```{r}
sports_owner_donations |>
  filter(str_detect(party, 'Democrat')) |>
  filter(election_year == 2020) |>
  summarise(sum(amount))
```
## 3: Which team owners gave the most donations?

    Donations from team owners, by cyle and party donated to: (this was in a table in the original story)
    
```{r}
# Who provided the highest number of donations? Charles Johnson, with 213 donations.
sports_owner_donations |>
  group_by(owner) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))

# Which league gave the most individual donations? MLB, with 746 donations.
sports_owner_donations |>
  group_by(league) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))
```

```{r}

```

## 4: "Overall, **51.7 percent of all contributions by WNBA owners went to Democrats**, with **42.3 percent going to Republicans.** And of the **\$1.3 million contributed by WNBA owners to Republicans, 65.5 percent comes from Kelly Loeffler,** the Republican from Georgia who is co-owner of the Atlanta Dream."

```{r}
sports_owner_donations |>
  filter(league == 'WNBA') |>
  filter(owner == 'Kelly Loeffler') |>
  summarise(wnba_amt = sum(amount)) |>
  arrange(desc(wnba_amt))

#Kelly Loeffler: $896,150

sports_owner_donations |>
  filter(league == 'WNBA') |>
  filter(str_detect(party, 'Republican')) |>
  summarise(sum(amount))
```
When adding up Republican WNBA spending, the total is $1,055,552 — this is not $1.3m, have to figure out why. We are not sure what the discrepancy is here, but we think it might be because there is not data for "Bipartisan, but mostly Democrat" included, which makes the total a little less than what FiveThirtyEight calculated. 

FiveThirtyEight had a searchable database, and we found that some of the donations that were there were not present in our data, particularly  Kelly Loeffler's bipartisan donations. Even when we tried to redownload the data, we got the same results, so this seems to be something with the data.

```{r}
sports_owner_donations |>
  filter(league == 'WNBA') |>
  group_by(party) |>
  summarise(total = sum(amount))
# there seems to be no WNBA donations for "Bipartisan, but mostly Democrat." 

sports_owner_donations |>
  filter(league == 'WNBA') |>
  group_by(owner) |>
  summarise(total = sum(amount)) |>
  arrange(desc(total))
```


## 5: "There are 40 owners who have donated at least \$100,000 to Republican causes over the six-year period, compared to 23 owners topping \$100,000 to Democrats, helping lead to these splits."

```{r}
# There are 37 rows of owners that donated more than $100,000 to Republican causes.
sports_owner_donations |>
  filter(str_detect(party, 'Republican')) |>
  group_by(owner) |>
  summarise(total = sum(amount)) |>
  filter(total >= 100000)

# 22 rows of owners that donated more than $100,000 to Democrat causes.
sports_owner_donations |>
  filter(str_detect(party, 'Democrat')) |>
  group_by(owner) |>
  summarise(total = sum(amount)) |>
  filter(total >= 100000)
```
These discrepancies are a little odd, as they are just slightly off from what FiveThirtyEight reported. We think that this could be because of the same issue with the WNBA calculations due to the data itself.

## Which year had the most donations?

A: 2018

```{r}
sports_owner_donations |>
  group_by(election_year) |>
  summarise(count_per_year = n()) |>
  arrange(desc(count_per_year))
```

## 6: "A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent **\$34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican** causes, compared to \$10.1 million (21.5 percent) to Democrats over the past three elections. Less than 6 percent of contributions went to bipartisan or unaffiliated recipients."

```{r}
# Money sent to purely Republican causes: 
# sum: 34,246,016 ~ $34.2 million, the total donation amount was $46,978,697
# 34,246,016/46,978,697 = 0.728969 = 72.9%
sports_owner_donations |>
  filter(party == 'Republican') |>
  summarise(sum(amount))

100*(34246016/46978697)

# Money sent to purely Democrat causes:
# sum: $10,113,639 ~ $10.1 million.
# $10,113,639/46,978,697 = 0.2152814 = 21.5%

sports_owner_donations |>
  filter(party == 'Democrat') |>
  summarise(sum(amount))

100*(10113639/46978697)
```


## Visualizing Data

Here, we created a column graph to represent the number of political donations by each league. Some labels include multiple leagues because there are owners who have a stake in multiple teams across various leagues.

```{r}
league_donations <- sports_owner_donations |>
  group_by(league) |>
  summarise(total_donation = n()) |>
  arrange(desc(total_donation))
```

```{r}
league_donations |>
  ggplot() +
  geom_bar(aes(x=league, weight=total_donation)) +
  labs(
    title="Number of Donations by League",
    x = "League",
    y = "Number of donations",
    caption = "source: ESPN and FiveThirtyEight"
  ) +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1))
```

## Final Memo

Reverse Engineering Deliverable 4 Memo

American professional sports owners have contributed nearly $47 million in federal elections since 2015. Owners contributed $10 million to Republican causes and $1.9 million to Democratic causes in the 2020 election cycle. Over the last three elections, 72.9 percent of contributions from sports owners went to Republican campaigns or super PACs and only 21.5 percent went to Democrats. This was a pattern throughout the study. The exception is the WNBA, in which 51.7 percent of contributions went to Democrats. The breakdown of contributions by league revealed that MLB owners have donated the most, and then the NBA, NHL, NFL, WNBA, and NASCAR. One of the most notable findings was that only a small group of donors made up a significant portion of the overall contribution data. This reflects on the power of these large companies with heavy public interest and endless funding to spend. The last and most obvious of the findings was that the biggest political spenders in professional sports ownership tended to support Republican candidates and causes.
Besides the data findings, the reporters included data visualizations in their final story. These visual representations of the data were effective in conceptualizing and comparing the very large numbers in donations. The story we looked at was just one in a series of six news stories FiveThirtyEight wrote with ESPN about political campaign donations in the sports industry. The data in this story sets the foundation for further investigations into the reasoning why sports owners are so secretive about their donation histories, and what motivates them to donate millions of dollars to campaigns in the first place. The actual names of the donors are published, not just the data about money, which allows readers to attach characters and real people to the figures about spending. If these names weren’t published, the story wouldn’t have as much of an impact. Since the owners are called out by name, they are held accountable for their donations, instead of getting away with covering up their political spending. 
The reporters address their main points in the first few paragraphs, stating that 160 owners across 125 professional teams were included in the research and state a few of the biggest data findings in those first few paragraphs. The reporters also use visuals such as graphs and tables to show their findings over the vast numbers in the dataset. The graphs are color coded in red and blue to represent Republican and Democrat, which makes them easy to understand. The story also includes a searchable database at the very bottom where readers can look up any specific sports owner or team to see the individual donations. This could be further improved if there was a way to see the total amounts donated by a given owner or team, as right now, a user would have to manually add up each donation in order to calculate the total.
The reporters mentioned that owners have “a number of ways to hide their political spending” and that they did not use contributions that could not be absolutely confirmed, even if it appeared to come from a particular owner. They also said that not all of the 2020 election cycle data was publicly available at the time of publishing, so because of these factors, their dataset doesn’t include all of the relevant data. Thus, we aren’t entirely sure how accurate or representative this data is for professional sports owners as a whole. It’s hard to tell how much relevant data is missing from this dataset; our numbers could look different if we had all of the information, which leaves us with questions.
The data came from the Federal Election Commision database of campaign finances for principal owners and Opensecrets. A lot of the data sets built themselves around existing data that they had from their sources. This means that a lot of research had been done prior to the study. It does not seem like the reporters actually spent a lot of time cleaning the data extensively before analysis. This is clear in the discrepancies found throughout the analysis. Still, the results are clear: most of the money was targeted toward the Republican cause. The numbers for the sports owners were very clearly presented in the data, which was certainly a benefit. However, the researchers just came to their own conclusions and built data sets off of it, which we would say is a weakness. These reporters seemingly took an easier route, and the Republican indication wasn’t much of a shock. Republicans favor tax cuts for the wealthy, all owners of sports teams are very wealthy. 

