---
title: "reverse_engineering_project"
author: "Hayden Sweeney"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

## Load libraries.

Load the required libraries for analyzing the data.

```{r}
options(scipen = 999)
library(tidyverse)
library(janitor)
library(readr)
library(lubridate)
```

## Loading and cleaning data

Load the data we will use for this analysis, and change the amount column from a character to a number.

```{r}
sports_owner_donations <- read_csv("sports-political-donations.csv") |>
  mutate(Amount = str_replace_all(Amount,'\\$','')) |>
  mutate(Amount = str_replace_all(Amount, '\\,','')) |>
  mutate(Amount = as.numeric(Amount))
```

In this notebook, we are reverse engineering what we believe to be the five most important findings from this story.

```{r}

```


## Q1: "Of the 160 owners and commissioners in ESPN’s database, 74 contributed a majority of their publicly searchable funds to Republican campaigns, with 48 donating a majority to Democrats and the rest sending most of their money to causes that aren’t fully aligned with either party."

## Who made the highest number of donations?
A1: Charles Johnson made the highest number of donations, with 213.

```{r}
sports_owner_donations |>
  group_by(Owner) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))
```

## Q2: ""Drilling down deeper into the data, we can see that federal contributions in each league lean red, except for in the WNBA."

## Which league made the most donations?
A1: The MLB made the most donations, with 746.

```{r}
sports_owner_donations |>
  group_by(League) |>
  summarise(total_donation = n()) |>
  arrange(desc(total_donation))
```

## Q3: "There are 40 owners who have donated at least $100,000 to Republican causes over the six-year period, compared to 23 owners topping $100,000 to Democrats"

## Which year had the highest number of donations?
A3: 2018 had the highest number of donations, with 997.

```{r}
sports_owner_donations |>
  group_by(`Election Year`) |>
  summarise(count_per_year = n()) |>
  arrange(desc(count_per_year))
```

