---
title: "reverse_engineering_project"
author: "Hayden Sweeney"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

## Load libraries.

Load the required libraries for analyzing the data.

```{r}
options(scipen = 999)
library(tidyverse)
library(janitor)
library(readr)
library(lubridate)
```

## Loading and cleaning data

Load the data we will use for this analysis, and change the amount column from a character to a number.

```{r}
sports_owner_donations <- read_csv("sports-political-donations.csv") |>
  mutate(Amount = str_replace_all(Amount,'\\$','')) |>
  mutate(Amount = str_replace_all(Amount, '\\,','')) |>
  mutate(Amount = as.numeric(Amount)) |>
  clean_names()
```

In this notebook, we are reverse engineering what we believe to be the five most important findings from this story.

## 1: ""Of the 160 owners and commissioners in ESPN’s database, 74 contributed a majority of their publicly searchable funds to Republican campaigns, with 48 donating a majority to Democrats and the rest sending most of their money to causes that aren’t fully aligned with either party."

## How many owners contributed donations to the Republican party?

Our answer is 104. This is different from the 74 number above likely because some of these owners onoy donated very little to the Republican party, and the majority to a different party or likely bipartisan. 

```{r}
sports_owner_donations |>
  filter(str_detect(party, 'Republican')) |>
  group_by(owner) |> 
  summarise(total = n())
```


## 2: "“American professional sports owners have contributed nearly $47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight, including $10 million to Republican causes and $1.9 million to Democratic causes so far in the 2020 election cycle.” 

## Find the total of all donations from all owners. 

The total amount that all owners have donated since 2015 is $46,978,697. 

```{r}
sports_owner_donations |>
  summarise(sum(amount))
```

## Find how much money owners donated to Republicans in the 2020 election cycle.

Owners donated $10,111,284 to Republicans in the 2020 election cycle.

```{r}
sports_owner_donations |>
  filter(str_detect(party, 'Republican')) |>
  filter(election_year == 2020) |>
  summarise(sum(amount))
```
## Find how much money owners donated to Democrats in the 2020 election cycle.

Owners donated $1,877,133 to Democrats in the 2020 election cycle.

```{r}
sports_owner_donations |>
  filter(str_detect(party, 'Democrat')) |>
  filter(election_year == 2020) |>
  summarise(sum(amount))
```
## 3: Which team owners gave the most donations?

    Donations from team owners, by cyle and party donated to: (this was in a table in the original story)
    
```{r}
# Who provided the highest number of donations? Charles Johnson, with 213 donations.
sports_owner_donations |>
  group_by(owner) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))

# Which league gave the most individual donations? MLB, with 746 donations.
sports_owner_donations |>
  group_by(league) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))
```

```{r}

```

## 4: "Overall, **51.7 percent of all contributions by WNBA owners went to Democrats**, with **42.3 percent going to Republicans.** And of the **\$1.3 million contributed by WNBA owners to Republicans, 65.5 percent comes from Kelly Loeffler,** the Republican from Georgia who is co-owner of the Atlanta Dream."

```{r}
sports_owner_donations |>
  filter(league == 'WNBA') |>
  filter(owner == 'Kelly Loeffler') |>
  summarise(wnba_amt = sum(amount)) |>
  arrange(desc(wnba_amt))

#Kelly Loeffler: $896,150

sports_owner_donations |>
  filter(league == 'WNBA') |>
  filter(str_detect(party, 'Republican')) |>
  summarise(sum(amount))
```
When adding up Republican WNBA spending, the total is $1,055,552 — this is not $1.3m, have to figure out why. We are not sure what the discrepancy is here, but we think it might be because there is not data for "Bipartisan, but mostly Democrat" included, which makes the total a little less than what FiveThirtyEight calculated. 

FiveThirtyEight had a searchable database, and we found that some of the donations that were there were not present in our data, particularly  Kelly Loeffler's bipartisan donations. Even when we tried to redownload the data, we got the same results, so this seems to be something with the data.

```{r}
sports_owner_donations |>
  filter(league == 'WNBA') |>
  group_by(party) |>
  summarise(total = sum(amount))
# there seems to be no WNBA donations for "Bipartisan, but mostly Democrat." 

sports_owner_donations |>
  filter(league == 'WNBA') |>
  group_by(owner) |>
  summarise(total = sum(amount)) |>
  arrange(desc(total))
```


## 5: "There are 40 owners who have donated at least \$100,000 to Republican causes over the six-year period, compared to 23 owners topping \$100,000 to Democrats, helping lead to these splits."

```{r}
# There are 37 rows of owners that donated more than $100,000 to Republican causes.
sports_owner_donations |>
  filter(str_detect(party, 'Republican')) |>
  group_by(owner) |>
  summarise(total = sum(amount)) |>
  filter(total >= 100000)

# 22 rows of owners that donated more than $100,000 to Democrat causes.
sports_owner_donations |>
  filter(str_detect(party, 'Democrat')) |>
  group_by(owner) |>
  summarise(total = sum(amount)) |>
  filter(total >= 100000)
```
These discrepancies are a little odd, as they are just slightly off from what FiveThirtyEight reported. We think that this could be because of the same issue with the WNBA calculations due to the data itself.

## Which year had the most donations?

A: 2018

```{r}
sports_owner_donations |>
  group_by(election_year) |>
  summarise(count_per_year = n()) |>
  arrange(desc(count_per_year))
```

## 6: "A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent **\$34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican** causes, compared to \$10.1 million (21.5 percent) to Democrats over the past three elections. Less than 6 percent of contributions went to bipartisan or unaffiliated recipients."

```{r}
# Money sent to purely Republican causes: 
# sum: 34,246,016 ~ $34.2 million, the total donation amount was $46,978,697
# 34,246,016/46,978,697 = 0.728969 = 72.9%
sports_owner_donations |>
  filter(party == 'Republican') |>
  summarise(sum(amount))

100*(34246016/46978697)

# Money sent to purely Democrat causes:
# sum: $10,113,639 ~ $10.1 million.
# $10,113,639/46,978,697 = 0.2152814 = 21.5%

sports_owner_donations |>
  filter(party == 'Democrat') |>
  summarise(sum(amount))

100*(10113639/46978697)
```


## Visualizing Data

Here, we created a column graph to represent the number of political donations by each league. Some labels include multiple leagues because there are owners who have a stake in multiple teams across various leagues.

```{r}
league_donations <- sports_owner_donations |>
  group_by(league) |>
  summarise(total_donation = n()) |>
  arrange(desc(total_donation))
```

```{r}
league_donations |>
  ggplot() +
  geom_bar(aes(x=league, weight=total_donation)) +
  labs(
    title="Number of Donations by League",
    x = "League",
    y = "Number of donations",
    caption = "source: ESPN and FiveThirtyEight"
  ) +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1))
```

