---
title: "reverse_engineering_project"
author: "Hayden Sweeney"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

## Load libraries.

Load the required libraries for analyzing the data.

```{r}
options(scipen = 999)
library(tidyverse)
library(janitor)
library(readr)
library(lubridate)
```

## Loading and cleaning data

Load the data we will use for this analysis, and change the amount column from a character to a number.

```{r}
sports_owner_donations <- read_csv("sports-political-donations.csv") |>
  mutate(Amount = str_replace_all(Amount,'\\$','')) |>
  mutate(Amount = str_replace_all(Amount, '\\,','')) |>
  mutate(Amount = as.numeric(Amount))
```

In this notebook, we are reverse engineering what we believe to be the five most important findings from this story.

## Q1: "“American professional sports owners have contributed nearly $47 million in federal elections since 2015, according to research by ESPN in partnership with FiveThirtyEight, including $10 million to Republican causes and $1.9 million to Democratic causes so far in the 2020 election cycle.”

## Find the total amount donated to Democrats in the 2020 election cycle.

```{r}
sports_owner_donations |>
  filter(str_detect(Party, 'Democrat')) |>
  filter(str_detect(Election Year, '2020')) |>
  total_amount = sum(Amount) |>
  arrange(desc(total_amount))
```


## Q2: "Of the 160 owners and commissioners in ESPN’s database, 74 contributed a majority of their publicly searchable funds to Republican campaigns, with 48 donating a majority to Democrats and the rest sending most of their money to causes that aren’t fully aligned with either party."

## Who made the highest number of donations?
Charles Johnson made the highest number of donations, with 213.

```{r}
sports_owner_donations |>
  group_by(Owner) |>
  summarise(total_donations = n()) |>
  arrange(desc(total_donations))
```


